Data Migration under a Changing Schema in Ampersand

### 2023-10-15 2022Migration/articleMigrationFACS.tex
WIP

%\def\id#1{\text{\it #1\/}}
\newcommand{\xrightarrowdbl}[2][]{%
  \xrightarrow[#1]{#2}\mathrel{\mkern-14mu}\rightarrow
}
\newcommand{\id}[1]{\text{\it #1\/}}
\newcommand{\code}[1]{\text{\tt\small #1}}
\newcommand{\stmtText}[1]{``{\small\tt #1}''}
\newcommand{\dom}[1]{\id{dom}(#1)}
\newcommand{\cod}[1]{\id{cod}(#1)}
%\renewcommand{\int}[2]{\id{inter}(#1,#2)}
\newcommand{\relsIn}[1]{\id{relsIn}(#1)}    % maps a Term to a set of Relations
\newcommand{\popF}[1]{\id{pop}_{#1}}
\newcommand{\pop}[2]{\popF{#1}(#2)}
\newcommand{\maintain}{\mathbin{\id{maint}}}
\newcommand{\enf}{\mathbin{{\tt enforce}}}
\newcommand{\enforce}[2]{{\tt enforce}_{#1}(#2)}
\newcommand{\instance}{\mathbin{\id{inst}}}
\newcommand{\relname}[1]{\id{relname}(#1)}
\newcommand{\evt}[2]{\id{event}_{#1,#2}}
\newcommand{\src}[1]{\id{src}(#1)}
\newcommand{\tgt}[1]{\id{tgt}(#1)}
\newcommand{\viol}[2]{\violC{#1}(#2)}
\newcommand{\violC}[1]{\id{viol}_{#1}}
\newcommand{\sign}[1]{\id{sign}(#1)}
\newcommand{\powerset}[1]{\cal{P}\{#1\}}
\newcommand{\la}{\langle}
\newcommand{\ra}{\rangle}
\newcommand{\full}{V}
\newcommand{\declare}[3]{\id{#1}_{\pair{#2}{#3}}}
\newcommand{\subst}[3]{#3_{[#1\rightarrow #2]}}
\newcommand{\fullt}[2]{V_{\pair{#1}{#2}}}
\newcommand{\iden}{I}
\newcommand{\ident}[1]{I_{\id{#1}}}
\newcommand{\expr}[3]{(#1)_{#2\times #3}}
\newcommand{\pair}[2]{\langle{#1},{#2}\rangle}
\newcommand{\maprel}[2]{{\tt maprel}_{#1}({#2})}
\newcommand{\Pair}[2]{#1\times#2}
\newcommand{\pairs}[1]{\id{pairs}(#1)}
\newcommand{\triple}[3]{\langle{#1},{#2},{#3}\rangle}
\newcommand{\quadruple}[4]{\langle{#1},{#2},{#3},{#4}\rangle}
\newcommand{\atom}[1]{{\tt\small #1}}
\newcommand{\atoms}{\mathcal{A}}
\newcommand{\Atoms}{\mathbb{A}}
%\newcommand{\events}{\mathcal{E}}
%\newcommand{\Events}{\mathbb{E}}
\newcommand{\concept}[1]{{\tt\small #1}}
\newcommand{\concepts}{\mathcal{C}}
\newcommand{\Concepts}{\mathbb{C}}
\newcommand{\decls}{\mathcal{D}}  %% names of relations
\newcommand{\rels}{\mathcal{R}}   %% all relations
\newcommand{\Rels}{\mathbb{R}}   %% all relations
\newcommand{\relations}{\mathcal{M}} % representing terms. M is a subset of R.
\newcommand{\triples}{\mathcal{T}}
\newcommand{\Triples}{\mathbb{T}}
\newcommand{\Triple}[3]{#1\times#2\times#3}
\newcommand{\vertices}{N}
\newcommand{\enforces}{\mathcal{E}}
\newcommand{\rules}{\mathcal{U}}
\newcommand{\Rules}{\mathbb{U}}
\newcommand{\specrules}{\mathcal{S}}
\newcommand{\roles}{\mathcal{O}}
\newcommand{\dataset}{\mathscr{D}}
\newcommand{\Dataset}{\mathbb{D}}
\newcommand{\schema}{\mathscr{Z}}
\newcommand{\functionality}{\mathscr{F}}
\newcommand{\select}[2]{\id{select}_{#1}\{{#2}\}}
\newcommand{\migrsys}{\mathscr{M}}
\newcommand{\infsys}{\mathscr{S}}
\newcommand{\tf}[1]{\mathscr{T}(#1)}
\newcommand{\ptf}[1]{\mathscr{T}'(#1)}
\newcommand{\ti}[1]{\mathscr{I}(#1)}
\newcommand{\tic}[1]{I_{\cal C}(#1)}
\newcommand{\relAdd}{\dagger}
\newcommand{\flip}[1]{{#1}^\smallsmile} %formerly:  {#1}^\backsim
\newcommand{\kleeneplus}[1]{{#1}^+}
\newcommand{\kleenestar}[1]{{#1}^*}
\newcommand{\cmpl}[1]{\overline{#1}}
\newcommand{\rel}{\times}
\newcommand{\compose}{;}
\newcommand{\subs}{\subseteq}%{\models}
\newcommand{\fun}{\rightarrow}
\newcommand{\isa}{\preceq}
%\newcommand{\isaClos}{\sqsubseteq}
\newcommand{\typetest}{?}
\newcommand{\meet}{\sqcap}
\newcommand{\join}{\sqcup}
\newcommand{\Meet}{\bigsqcap}
\newcommand{\Moin}{\bigsqcup} % because LaTeX has already defined command \Join.
\newcommand{\order}{\ominus}
\newcommand{\anything}{\top}
\newcommand{\nothing}{\bot}
\newcommand{\rewriteto}{\rightarrow}
\newcommand{\calc}{\implies}
\newcommand{\alland}{\bigwedge}
\newcommand{\mph}[3]{#1_{#2\times #3}}
\newcommand{\mphu}[1]{#1_{\univ\times\univ}}

@@

   \begin{definition}[Data set]
   A data set $\dataset$ is a tuple $\pair{\triples}{\instance}$ with finite $\triples \subseteq {\Triple{\Atoms}{\Rels}{\Atoms}}$ and $\instance \subseteq  {\Pair{\Atoms}{\Concepts}}$ that satisfies:
\begin{eqnarray}
   \triple{a}{\declare{n}{A}{B}}{b}\in\triples&\Rightarrow&a\instance A\ \wedge\ b\instance B
   \label{eqn:wellTypedEdge}
\end{eqnarray}
\end{definition}
   Looking at the example,
   equation~\ref{eqn:wellTypedEdge} says that \atom{Peter} is an instance of {\tt Person} and \atom{074238991} is an instance of {\tt TelephoneNumber}.
   In practice, users can say that the Person Peter has telephone number 074238991.
   So, the ``thing'' that \atom{Peter} refers to (which is Peter) has \atom{074238991} as a telephone number.
   The notations $\triples_{\dataset}$ and $\instance_{\dataset}$ are used to disambiguate $\triples$ and $\instance$ when necessary.
   To save writing in the sequel, the notation $a\ r\ b$ means that $\triple{a}{r}{b}\in\triples$.

@@

\subsection{Violations}
\label{sct:Violations}
   To satisfy a constraint $u$ means that it is true for all triples in the data set.
   Let us assume that we can represent each violation of a constraint $u$ as a pair
   and let the function $\violC{u}:\Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$ produce these violations.
   Note that $\viol{u}{\dataset}=\emptyset$ means that $\dataset$ satisfies $u$.
   Now we can represent an invariant $u$ as a triple $\triple{\violC{u}}{A}{B}$.
   It must be well-typed, just like a relation:
\begin{equation}
   \pair{a}{b}\in\viol{\triple{\violC{u}}{A}{B}}{\dataset}\ \Rightarrow\ a\instance A\ \wedge\ b\instance B
\end{equation}
   We represent an enforced invariant $u$ as a pair $\pair{\declare{n}{A}{B}}{\violC{u}}$,
   where relation $\declare{n}{A}{B}$ is used to restore invariance of the enforced invariant.
   The enforced invariant must be well-typed:
\begin{equation}
   \pair{a}{b}\in\viol{\pair{\declare{n}{A}{B}}{\violC{u}}}{\dataset}\ \Rightarrow\ a\instance A\ \wedge\ b\instance B
\end{equation}
   Besides, an enforced invariant must be automatically restorable by adding the violations to relation $\declare{n}{A}{B}$:
\begin{equation}
   \viol{u}{\dataset\cup\{\triple{a}{\declare{n}{A}{B}}{b}\mid\pair{a}{b}\in\viol{u}{\dataset}\}}=\emptyset
\end{equation}
   (TODO: define $\cup$ for data sets.)
   The language Ampersand has more types of enforced invariants than just this one,
   but to define this one only is sufficient for this paper.

>>>>

%\def\id#1{\text{\it #1\/}}
\newcommand{\xrightarrowdbl}[2][]{%
  \xrightarrow[#1]{#2}\mathrel{\mkern-14mu}\rightarrow
}
\newcommand{\id}[1]{\text{\it #1\/}}
\newcommand{\code}[1]{\text{\tt\small #1}}
\newcommand{\stmtText}[1]{``{\small\tt #1}''}
\newcommand{\dom}[1]{\id{dom}(#1)}
\newcommand{\cod}[1]{\id{cod}(#1)}
%\renewcommand{\int}[2]{\id{inter}(#1,#2)}
\newcommand{\relsIn}[1]{\id{relsIn}(#1)}    % maps a Term to a set of Relations
\newcommand{\popF}[1]{\id{pop}_{#1}}
\newcommand{\pop}[2]{\popF{#1}(#2)}
\newcommand{\maintain}{\mathbin{\id{maint}}}
\newcommand{\enf}{\mathbin{{\tt enforce}}}
\newcommand{\enforce}[2]{{\tt enforce}_{#1}(#2)}
\newcommand{\instance}{\mathbin{\id{inst}}}
\newcommand{\relname}[1]{\id{relname}(#1)}
\newcommand{\evt}[2]{\id{event}_{#1,#2}}
\newcommand{\src}[1]{\id{src}(#1)}
\newcommand{\tgt}[1]{\id{tgt}(#1)}
\newcommand{\viol}[2]{\violC{#1}(#2)}
\newcommand{\violC}[1]{\id{viol}_{#1}}
\newcommand{\sign}[1]{\id{sign}_{#1}}
\newcommand{\enfRel}[1]{\id{enf}_{#1}}
\newcommand{\signature}[2]{\langle{#1},{#2}\rangle}
\newcommand{\powerset}[1]{\cal{P}\{#1\}}
\newcommand{\la}{\langle}
\newcommand{\ra}{\rangle}
\newcommand{\full}{V}
\newcommand{\declare}[3]{\id{#1}_{\pair{#2}{#3}}}
\newcommand{\subst}[3]{#3_{[#1\rightarrow #2]}}
\newcommand{\fullt}[2]{V_{\pair{#1}{#2}}}
\newcommand{\iden}{I}
\newcommand{\ident}[1]{I_{\id{#1}}}
\newcommand{\expr}[3]{(#1)_{#2\times #3}}
\newcommand{\pair}[2]{\langle{#1},{#2}\rangle}
\newcommand{\maprel}[2]{{\tt maprel}_{#1}({#2})}
\newcommand{\Pair}[2]{#1\times#2}
\newcommand{\pairs}[1]{\id{pairs}(#1)}
\newcommand{\triple}[3]{\langle{#1},{#2},{#3}\rangle}
\newcommand{\quadruple}[4]{\langle{#1},{#2},{#3},{#4}\rangle}
\newcommand{\atom}[1]{{\tt\small #1}}
\newcommand{\atoms}{\mathcal{A}}
\newcommand{\Atoms}{\mathbb{A}}
%\newcommand{\events}{\mathcal{E}}
%\newcommand{\Events}{\mathbb{E}}
\newcommand{\concept}[1]{{\tt\small #1}}
\newcommand{\concepts}{\mathcal{C}}
\newcommand{\Concepts}{\mathbb{C}}
\newcommand{\decls}{\mathcal{D}}  %% names of relations
\newcommand{\rels}{\mathcal{R}}   %% all relations
\newcommand{\Rels}{\mathbb{R}}   %% all relations
\newcommand{\relations}{\mathcal{M}} % representing terms. M is a subset of R.
\newcommand{\triples}{\mathcal{T}}
\newcommand{\Triples}{\mathbb{T}}
\newcommand{\Triple}[3]{#1\times#2\times#3}
\newcommand{\vertices}{N}
\newcommand{\enforces}{\mathcal{E}}
\newcommand{\rules}{\mathcal{U}}
\newcommand{\Rules}{\mathbb{U}}
\newcommand{\specrules}{\mathcal{S}}
\newcommand{\roles}{\mathcal{O}}
\newcommand{\dataset}{\mathscr{D}}
\newcommand{\Dataset}{\mathbb{D}}
\newcommand{\schema}{\mathscr{Z}}
\newcommand{\functionality}{\mathscr{F}}
\newcommand{\select}[2]{\id{select}_{#1}\{{#2}\}}
\newcommand{\migrsys}{\mathscr{M}}
\newcommand{\infsys}{\mathscr{S}}
\newcommand{\tf}[1]{\mathscr{T}(#1)}
\newcommand{\ptf}[1]{\mathscr{T}'(#1)}
\newcommand{\ti}[1]{\mathscr{I}(#1)}
\newcommand{\tic}[1]{I_{\cal C}(#1)}
\newcommand{\relAdd}{\dagger}
\newcommand{\flip}[1]{{#1}^\smallsmile} %formerly:  {#1}^\backsim
\newcommand{\kleeneplus}[1]{{#1}^+}
\newcommand{\kleenestar}[1]{{#1}^*}
\newcommand{\cmpl}[1]{\overline{#1}}
\newcommand{\rel}{\times}
\newcommand{\compose}{;}
\newcommand{\subs}{\subseteq}%{\models}
\newcommand{\fun}{\rightarrow}
\newcommand{\isa}{\preceq}
%\newcommand{\isaClos}{\sqsubseteq}
\newcommand{\typetest}{?}
\newcommand{\meet}{\sqcap}
\newcommand{\join}{\sqcup}
\newcommand{\Meet}{\bigsqcap}
\newcommand{\Moin}{\bigsqcup} % because LaTeX has already defined command \Join.
\newcommand{\order}{\ominus}
\newcommand{\anything}{\top}
\newcommand{\nothing}{\bot}
\newcommand{\rewriteto}{\rightarrow}
\newcommand{\calc}{\implies}
\newcommand{\alland}{\bigwedge}
\newcommand{\mph}[3]{#1_{#2\times #3}}
\newcommand{\mphu}[1]{#1_{\univ\times\univ}}

@@

   \begin{definition}[Data set]
   A data set $\dataset$ is a tuple $\pair{\triples}{\instance}$ with finite $\triples \subseteq {\Triple{\Atoms}{\Rels}{\Atoms}}$ and $\instance\subseteq\Pair{\Atoms}{\Concepts}$ that satisfies:
\begin{eqnarray}
   \triple{a}{\declare{n}{A}{B}}{b}\in\triples&\Rightarrow&a\instance A\ \wedge\ b\instance B
   \label{eqn:wellTypedEdge}
\end{eqnarray}
\end{definition}
   Looking at the example,
   equation~\ref{eqn:wellTypedEdge} says that \atom{Peter} is an instance of {\tt Person} and \atom{074238991} is an instance of {\tt TelephoneNumber}.
   In practice, users can say that the Person Peter has telephone number 074238991.
   So, the ``thing'' that \atom{Peter} refers to (which is Peter) has \atom{074238991} as a telephone number.
   The notations $\triples_{\dataset}$ and $\instance_{\dataset}$ are used to disambiguate $\triples$ and $\instance$ when necessary.
   To save writing in the sequel, the notation $a\ r\ b$ means that $\triple{a}{r}{b}\in\triples$.

@@

\subsection{Violations}
\label{sct:Violations}
   To satisfy a constraint $u$ means that it is true for all triples in the data set.
   Let us assume that we can represent each violation of a constraint $u$ as a pair.
   Let the function $\violC{u}:\Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$ produce these violations
   and let the function $\sign{u}:\Pair{\Concepts}{\Concepts}$ tell their signature:
\begin{equation}
   \pair{a}{b}\in\viol{u}{\dataset}\ \wedge\ \sign{u}=\pair{A}{B}\Rightarrow\ a\instance A\ \wedge\ b\instance B
\end{equation}
   Note that $\viol{u}{\dataset}=\emptyset$ means that $\dataset$ satisfies $u$.

   In case $u$ is an enforced invariant,
   the system will keep it satisfied by adding the violations to relation $\declare{n}{A}{B}$:
\begin{equation}
   \viol{u}{\pair{\triples\cup\{\triple{a}{\declare{n}{A}{B}}{b}\mid\pair{a}{b}\in\viol{u}{\dataset}\}}{\instance\cup\instance'}}=\emptyset
\end{equation}
   Let us call $\declare{n}{A}{B}$ the {enforced relation} of $u$, defined by $\enfRel$:
\begin{equation}
   \enfRel{u}=\declare{n}{A}{B}
\end{equation}
   The language Ampersand has more types of enforced invariants than just this one,
   but to define this one only is sufficient for this paper.
